A Linear Solution of the Linearity Challenge

### 2025-12-13 paper/Reduction.lagda.tex
restored reduction

\begin{code}[hide]
{-# OPTIONS --rewriting #-}
open import Data.Bool using (Bool; true; false; if_then_else_)
open import Data.Product using (_×_; _,_; ∃; Σ; Σ-syntax; ∃-syntax)
open import Relation.Binary.PropositionalEquality using (_≡_; refl; sym; cong; cong₂)
open import Data.List.Base using (List; []; _∷_; [_]; _++_)
open import Data.List.Properties using (++-assoc)

@@

\begin{code}[hide]
weakening : ∀{Γ Γ₁ Γ₂} (un : Un Γ₁) → Γ ≃ Γ₁ + Γ₂ → Process Γ₂ → Process Γ
weakening un p P = ↭process (↭concat p) (aux un P)
  where
    aux : ∀{Γ₁ Γ₂} (un : Un Γ₁) → Process Γ₂ → Process (Γ₁ ++ Γ₂)
    aux un-[] P = P
    aux (un-∷ un) P = weaken (< ≫) (aux un P)

contraction : ∀{Γ Γ₁ Γ₂} (un : Un Γ₁) → Γ ≃ Γ₁ + Γ₂ → Process (Γ₁ ++ Γ) → Process Γ
contraction un p P = ↭process (↭concat p) (aux un (↭process (↭left (↭sym (↭concat p))) P))
  where
    aux : ∀{Γ₁ Γ₂} → Un Γ₁ → Process (Γ₁ ++ Γ₁ ++ Γ₂) → Process (Γ₁ ++ Γ₂)
    aux un-[] P = P
    aux {`? A ∷ Γ₁} {Γ₂} (un-∷ un) P with contract (< ≫) (↭process (↭shift {`? A} {`? A ∷ Γ₁} {Γ₁ ++ Γ₂}) P)
    ... | P₁ rewrite sym (++-assoc (`? A ∷ Γ₁) Γ₁ Γ₂) with ↭process (↭sym (↭shift {`? A} {Γ₁ ++ Γ₁})) P₁
    ... | P₂ rewrite ++-assoc Γ₁ Γ₁ (`? A ∷ Γ₂) with aux un P₂
    ... | P₃ = ↭process ↭shift P₃
\end{code}

@@

\begin{code}
data _↝_ {Γ} : Process Γ → Process Γ → Set where
\end{code}

@@

\begin{code}
  r-link      : ∀{Δ A P} (p : Γ ≃ [ dual A ] + Δ) →
                cut {A} p (link (< > •)) P ↝ ↭process (↭concat p) P
\end{code}

The splitting $p$ indicates that $y$ (of type $\dual{A}$) occurs on the left
sub-process of the cut (that is the link $\Link\x\y$) and the splitting
$(\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{>}~\AgdaInductiveConstructor{•})$
to which \AgdaInductiveConstructor{link} is applied is structured consistently
with the syntax of the link $\Link\x\y$ being reduced, which is oriented so that
the restricted channel $x$ occurs on the left. 
%
The process $P$ has type
$\AgdaDatatype{Process}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\ContextD)$,
and turns into $P\subst\y\x$ after the reduction. The type
$\AgdaFunction{dual}~A$ of $x$ in $P$ is the first in the typing context,
indicating that it is the newest channel that $P$ is aware of. However, after
the reduction, $x$ is replaced by $y$ which is found \emph{somewhere} within
$\Context$. The exact location of $y$ in $\ContextC$ is found in the splitting
$p$. We ``rename'' $x$ into $y$ within $P$ using the permutation
$\AgdaFunction{↭concat}~p$.

@@

\begin{code}
  r-close     : ∀{P} (p₀ : Γ ≃ [] + Γ) (q₀ : Γ ≃ [] + Γ) →
                cut p₀ close (wait (< q₀) P) ↝ P
\end{code}

@@

The constructor \AgdaInductiveConstructor{r-left} models \RSelect when the
selected tag is $\InTag_1$:

\begin{code}
  r-left      : ∀{Γ₁ Γ₂ A B P Q R}
                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
                cut {A ⊕ B} p (left (< p₀) P)
                              (case (< q₀) Q R) ↝ cut p P Q
\end{code}

There is not much more to note here except again for the multiple
quantifications over the trivial splittings $\Context_i \simeq
\AgdaInductiveConstructor{[]} + \Context_i$ and the use of
\AgdaInductiveConstructor{<} to make sure that the channel referred to by
\AgdaInductiveConstructor{left} and \AgdaInductiveConstructor{case} is indeed
the one restricted by the cut.

\begin{code}[hide]
  r-right     : ∀{Γ₁ Γ₂ A B P Q R}
                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
                cut {A ⊕ B} p (right (< p₀) P)
                              (case (< q₀) Q R) ↝ cut p P R
  r-fork      : ∀{Γ₁ Γ₂ Γ₃ Δ A B P Q R}
                (p : Γ ≃ Δ + Γ₃) (p₀ : Γ₃ ≃ [] + Γ₃) (q : Δ ≃ Γ₁ + Γ₂) (q₀ : Δ ≃ [] + Δ) →
                let _ , p′ , q′ = +-assoc-l p q in
                cut {A ⊗ B} p (fork (< q₀) q P Q)
                              (join (< p₀) R) ↝ cut q′ P (cut (> p′) Q R)
  r-client    : ∀{Γ₁ Γ₂ A P Q}
                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
                cut {`! A} p (server (< p₀) un P) (client (< q₀) Q) ↝ cut p P Q
  r-weaken    : ∀{Γ₁ Γ₂ A P Q}
                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
                cut {`! A} p (server (< p₀) un P)
                             (weaken (< q₀) Q) ↝ weakening un p Q
  r-contract  : ∀{Γ₁ Γ₂ A P Q}
                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₁) →
                cut {`! A} p (server (< p₀) un P)
                             (contract (< q₀) Q) ↝
                    contraction un p
                      (cut ++≃+ (server (< p₀) un P)
                      (cut (> p) (server (< p₀) un P) Q))
\end{code}

The remaining constructors that describe the base reductions follow a similar
pattern, noting that the implementation of \RWeaken and \RContract makes use of
auxiliary functions that respectively weaken and contract the typing context of
the resulting process as shown in \Cref{tab:semantics}. It is worth glancing at
\RExists since it involves a non-trivial rewriting of types:

\begin{code}
  r-exists     : ∀{A B Γ₁ Γ₂ P F}
                 (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
                 cut {`∃ A} p (ex {_} {B} (< p₀) P) (all (< q₀) F) ↝ cut p P (F B)
\end{code}

Recalling the definitions of \AgdaInductiveConstructor{ex} and
\AgdaInductiveConstructor{all} from \Cref{sec:process-agda}, we note that $P$
has type
$\AgdaDatatype{Process}~(\AgdaFunction{subst}~[~B~/\_]~A~\AgdaInductiveConstructor{∷}~\Context_1)$
and $(F~B)$ is a process of type
$\AgdaDatatype{Process}~(\AgdaFunction{subst}~[~B~/\_]~(\AgdaFunction{dual}~A)~\AgdaInductiveConstructor{∷}~\Context_2)$.
In order for these two process to be composable in a cut, it must be the case
that $\AgdaFunction{subst}~[~B~/\_]~(\AgdaFunction{dual}~A) \equiv
\AgdaFunction{dual}~(\AgdaFunction{subst}~[~B~/\_]~A)$, which was proved in
\Cref{sec:type-agda} under the name \AgdaFunction{dual-subst}. Thanks to the
implicit rewriting rule, we do not have to rewrite the index in the type of
$(F~B)$, which is silently accepted as is.

@@

\begin{code}
  r-cut        : ∀{Γ₁ Γ₂ A P Q R} (q : Γ ≃ Γ₁ + Γ₂) →
                 P ↝ Q → cut {A} q P R ↝ cut q Q R
  r-cong       : ∀{P R Q} → P ⊒ R → R ↝ Q → P ↝ Q
\end{code}

>>>>

\begin{code}[hide]
{-# OPTIONS --rewriting #-}
open import Data.Sum using (inj₁; inj₂)
open import Data.Product using (_,_)
open import Data.List.Base using ([]; _∷_; [_]; _++_)
open import Data.List.Properties using (++-assoc)
open import Relation.Unary
open import Relation.Binary.PropositionalEquality using (sym)

@@

\begin{code}[hide]
weakening : ∀[ Un ∗ Proc ⇒ Proc ]
weakening (un ⟨ p ⟩ P) = ↭proc (↭concat p) (aux un P)
  where
    aux : ∀{Γ₁ Γ₂} (un : Un Γ₁) → Proc Γ₂ → Proc (Γ₁ ++ Γ₂)
    aux un-[] P = P
    aux (un-∷ un) P = weaken (ch ⟨ < ≫ ⟩ aux un P)

contraction : ∀{Γ Γ₁ Γ₂} (un : Un Γ₁) → Γ ≃ Γ₁ + Γ₂ → Proc (Γ₁ ++ Γ) → Proc Γ
contraction un p P = ↭proc (↭concat p) (aux un (↭proc (↭left (↭sym (↭concat p))) P))
  where
    aux : ∀{Γ₁ Γ₂} → Un Γ₁ → Proc (Γ₁ ++ Γ₁ ++ Γ₂) → Proc (Γ₁ ++ Γ₂)
    aux un-[] P = P
    aux {`? A ∷ Γ₁} {Γ₂} (un-∷ un) P with contract (ch ⟨ < ≫ ⟩ (↭proc (↭shift {`? A} {`? A ∷ Γ₁} {Γ₁ ++ Γ₂}) P))
    ... | P₁ rewrite sym (++-assoc (`? A ∷ Γ₁) Γ₁ Γ₂) with ↭proc (↭sym (↭shift {`? A} {Γ₁ ++ Γ₁})) P₁
    ... | P₂ rewrite ++-assoc Γ₁ Γ₁ (`? A ∷ Γ₂) with aux un P₂
    ... | P₃ = ↭proc ↭shift P₃
\end{code}

@@

\begin{code}
data _↝_ {Γ} : Proc Γ → Proc Γ → Set where
\end{code}

@@

\begin{code}
  r-link      : ∀{Δ A P} (p : Γ ≃ [ dual A ] + Δ) →
                cut {A} (link (ch ⟨ < > • ⟩ ch) ⟨ p ⟩ P) ↝ ↭proc (↭concat p) P
\end{code}

The splitting $p$ indicates that $y$ (of type $\dual{A}$) occurs on the left
sub-process of the cut (that is the link $\Link\x\y$) and the splitting
$(\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{>}~\AgdaInductiveConstructor{•})$
to which \AgdaInductiveConstructor{link} is applied is structured consistently
with the syntax of the link $\Link\x\y$ being reduced, which is oriented so that
the restricted channel $x$ occurs on the left. 
%
The process $P$ has type
$\AgdaDatatype{Proc}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\ContextD)$,
and turns into $P\subst\y\x$ after the reduction. The type
$\AgdaFunction{dual}~A$ of $x$ in $P$ is the first in the typing context,
indicating that it is the newest channel that $P$ is aware of. However, after
the reduction, $x$ is replaced by $y$ which is found \emph{somewhere} within
$\Context$. The exact location of $y$ in $\ContextC$ is found in the splitting
$p$. We ``rename'' $x$ into $y$ within $P$ using the permutation
$\AgdaFunction{↭concat}~p$.

@@

\begin{code}
  r-close     : ∀{P} (p₀ q₀ : Γ ≃ [] + Γ) →
                cut (close ch ⟨ p₀ ⟩ wait (ch ⟨ < q₀ ⟩ P)) ↝ P
\end{code}

@@

The constructor \AgdaInductiveConstructor{r-select-l} models \RSelect when the
selected tag is $\InTag_1$:

\begin{code}
  r-select-l  : ∀{Γ₁ Γ₂ A B P Q R}
                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
                cut {A ⊕ B} (select (ch ⟨ < p₀ ⟩ inj₁ P) ⟨ p ⟩
                             case (ch ⟨ < q₀ ⟩ (Q , R))) ↝ cut (P ⟨ p ⟩ Q)
\end{code}

There is not much to note here except again for the multiple quantifications
over the trivial splittings $\Context_i \simeq \AgdaInductiveConstructor{[]} +
\Context_i$ and the use of \AgdaInductiveConstructor{<} to make sure that the
channel referred to by \AgdaInductiveConstructor{select} and
\AgdaInductiveConstructor{case} is indeed the one restricted by the cut.

\begin{code}[hide]
  r-select-r  : ∀{Γ₁ Γ₂ A B P Q R}
                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
                cut {A ⊕ B} (select (ch ⟨ < p₀ ⟩ inj₂ P) ⟨ p ⟩
                             case (ch ⟨ < q₀ ⟩ (Q , R))) ↝ cut (P ⟨ p ⟩ R)
  r-fork      : ∀{Γ₁ Γ₂ Γ₃ Δ A B P Q R}
                (p : Γ ≃ Δ + Γ₃) (p₀ : Γ₃ ≃ [] + Γ₃) (q : Δ ≃ Γ₁ + Γ₂) (q₀ : Δ ≃ [] + Δ) →
                let _ , p′ , q′ = +-assoc-l p q in
                cut {A ⊗ B} (fork (ch ⟨ < q₀ ⟩ (P ⟨ q ⟩ Q)) ⟨ p ⟩
                             join (ch ⟨ < p₀ ⟩ R)) ↝ cut (P ⟨ q′ ⟩ cut (Q ⟨ > p′ ⟩ R))
  r-client    : ∀{Γ₁ Γ₂ A P Q}
                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₂) →
                cut {`? A} (client (ch ⟨ < p₀ ⟩ P) ⟨ p ⟩ server (ch ⟨ < q₀ ⟩ (un , Q))) ↝
                cut (P ⟨ p ⟩ Q)
  r-weaken    : ∀{Γ₁ Γ₂ A P Q}
                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₂) →
                cut {`? A} (weaken (ch ⟨ < p₀ ⟩ P) ⟨ p ⟩ server (ch ⟨ < q₀ ⟩ (un , Q))) ↝
                weakening (un ⟨ +-comm p ⟩ P)
  r-contract  : ∀{Γ₁ Γ₂ A P Q}
                (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) (un : Un Γ₂) →
                cut {`? A} (contract (ch ⟨ < p₀ ⟩ P) ⟨ p ⟩ server (ch ⟨ < q₀ ⟩ (un , Q))) ↝
                contraction un (+-comm p) (cut (server (ch ⟨ < q₀ ⟩ (un , Q)) ⟨ ++≃+ ⟩
                                           cut (server (ch ⟨ < q₀ ⟩ (un , Q)) ⟨ > +-comm p ⟩ P)))
\end{code}

The remaining constructors that describe the base reductions follow a similar
pattern, noting that the implementation of \RWeaken and \RContract makes use of
auxiliary functions that respectively weaken and contract the typing context of
the resulting process as shown in \Cref{tab:semantics}. It is worth glancing at
the implementation of \RExists since it involves a non-trivial rewriting of
types:

\begin{code}
  r-exists     : ∀{A B Γ₁ Γ₂ P F}
                 (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
                 cut {`∃ A} (ex {A} {B} (ch ⟨ < p₀ ⟩ P) ⟨ p ⟩ all (ch ⟨ < q₀ ⟩ F)) ↝
                 cut (P ⟨ p ⟩ F B)
\end{code}

Recalling the definitions of \AgdaInductiveConstructor{ex} and
\AgdaInductiveConstructor{all} from \Cref{sec:process-agda}, we note that $P$
has type
$\AgdaDatatype{Proc}~(\AgdaFunction{subst}~[~B~/]~A~\AgdaInductiveConstructor{∷}~\Context_1)$
and $(F~B)$ is a process of type
$\AgdaDatatype{Proc}~(\AgdaFunction{subst}~[~B~/]~(\AgdaFunction{dual}~A)~\AgdaInductiveConstructor{∷}~\Context_2)$.
In order for these two process to be composable in a cut, it must be the case
that $\AgdaFunction{subst}~[~B~/]~(\AgdaFunction{dual}~A) \equiv
\AgdaFunction{dual}~(\AgdaFunction{subst}~[~B~/]~A)$, which was proved in
\Cref{sec:type-agda} under the name \AgdaFunction{dual-subst}. Thanks to the
implicit rewriting rule, we do not have to rewrite the index in the type of
$(F~B)$, which is silently accepted as is.

@@

\begin{code}
  r-cut        : ∀{Γ₁ Γ₂ A P Q R} (q : Γ ≃ Γ₁ + Γ₂) →
                 P ↝ Q → cut {A} (P ⟨ q ⟩ R) ↝ cut (Q ⟨ q ⟩ R)
  r-cong       : ∀{P R Q} → P ⊒ R → R ↝ Q → P ↝ Q
\end{code}
